---
title: "Creating samples with sampler"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Creating samples with sampler}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(whSample)
```

The ***sampler*** function is part of the ***whSample*** package of utilities for sampling. It is a menu-driven R application that reads an Excel spreadsheet or delimited text file and generates a new Excel workbook with a copy of the original data, a Simple Random or Stratified Random Sample, and a report of what it did.

The *sampler* function calls [***ssize***](Using_ssize.html) to get its sample size estimate. *Sampler* uses the same sample size defaults as *ssize* and they can be overridden when launching *sampler* from the command line.

***Sampler*** takes two additional arguments:

- **backups** provides a buffer for use if necessary to replace samples found to be invalid for some reason, and
- **seed**, used to seed the internal random number generator.

The full range of command-line options is:

```{r eval=FALSE}
sampler(ci=0.95, me=0.07, p=0.5, backups=5, seed=NULL)
```


The defaults for these additional arguments are *backups=5* and *seed=NULL*. The default seed will tell *sampler* to use the current system time in milliseconds. Any number can be used as a seed. Whichever one is used will be listed in the *Report* output tab.

To override any of these defaults, enter *name=value* as an argument when calling *sampler*, as in:

```{r eval=F}
sampler(p=0.06)
```

. More on changing the number of backups in a minute.

## Running ***sampler***

The easiest way to run *sampler* is by calling it from the R console:

```{r eval=FALSE}
sampler()
```

### Choose the source data file

A file chooser will appear to select a source file. *Sampler* will remember where the source file came from and will put the resulting output file in the same folder with the name ***source file name*** **Sample.xlsx**.

*Sampler* will read Excel or delimited text files.

### Choose the data worksheet

If the source file is in an Excel workbook, *sampler* will display a list of worksheets in that file for you to choose from.

### Choose the number of backups

A "Number of Backups" menu will offer either zero, five, or ten backups (per stratum). The default (if you just hit *Okay*) is five.

There may be times when you want a specific number of backups that isn't listed. In this case, you need to **start** ***sampler*** **with the number of backups desired**, as in:

```{r eval=FALSE}
sampler(backups=25)
```

**Important note:** At this point, if you specified a number of backups on the command line, **hit** ***Cancel*** and *sampler* will use the number you specified. Clicking Okay will accept whatever was highlighted in the menu. If you didn't specify a number on the command line and hit Cancel, *sampler* will use the default of five.

For a Simple Random Sample, the startup command shown above will add the next 25 items from the randomized population to a separate table at the end of the output Excel workbook.

Stratified samples will include 25 backups **for each stratum**. This could be a problem if a given stratum doesn't have 25 more items to include. In this case, *sampler* will scale back the number of backups to the number of items available for that stratum. The ***Report tab*** will list the number of backups actually provided for each stratum.

### Choose the sampling type

*Sampler* can create a Simple Random Sample, a Stratified Random Sample, or a variation of the stratified sample where each stratum is listed with its backups in a separate Excel worksheet. A "Sample Type" menu provides the options.

### Choose what to stratify on

The "Stratify on" menu lists all the column headers in the selected worksheet. *Sampler* will group all the samples by this name and split out the results accordingly.

## Output

*Sampler* creates a new Excel workbook in three parts:

  - a copy of the original (source) data,
  - an Excel spreadsheet with the requested sample, and
  - a new tab called *Report* with key reference information:

    - path and name of the source file
    - size (in rows) of the source file
    - sample type (Simple Random Sample, Stratified Random Sample, or Tabbed Stratified Sample)
    - sample size
    - number of strata
    - number of backups per stratum
    - random number seed used, for documentation and reproducibility
    - date-time stamp of when the sample was generated
    - stratification information (stratified samples only):
      
      - name
      - size of the population
      - proportion of the population for each stratum
      - the number of primary samples
      - the actual number of backup samples (may not be the same as that requested)

## Installation

To install from CRAN, use:

```{r eval=FALSE}
install.packages("whSample")
```

All necessary dependencies also will be installed as necessary.

You can install\* the latest development version of whSample from the R console with:

``` {r eval=FALSE}
devtools::install_github("km4ivi/whSample")
```
\* This requires the `devtools` package to be installed first.

### Other necessary packages

*sampler* depends on several external packages to run properly. Ensure these packages are installed:

- tidyverse (or individually: magrittr, dplyr, purrr)
- openxlsx
- data.table
- tools
- glue

## Examples

*sampler()*: Uses all defaults, gets N from the source data.

*sampler(backups=10, seed=12345)*: Overrides specific defaults
